---
title: "GDPx4 DRUG-seq Full Analysis (Rmd + Reticulate + Polars)"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  max_rows: 0
  top_n: 50
  top_compounds_heatmap: 10
  top_genes_heatmap: 30
  target_compound: ""
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

suppressPackageStartupMessages({
  library(ggplot2)
  library(reticulate)
})

root_candidates <- c(
  normalizePath(getwd(), winslash = "/", mustWork = TRUE),
  normalizePath(file.path(getwd(), "..", ".."), winslash = "/", mustWork = FALSE)
)
project_root <- root_candidates[file.exists(file.path(root_candidates, ".git"))][1]
if (is.na(project_root) || !nzchar(project_root)) {
  project_root <- normalizePath(getwd(), winslash = "/", mustWork = TRUE)
}

venv_candidates <- c(
  file.path(project_root, ".venv-gdpx4", "bin", "python"),
  file.path(getwd(), ".venv-gdpx4", "bin", "python")
)
venv_python <- venv_candidates[file.exists(venv_candidates)][1]
if (is.na(venv_python) || !file.exists(venv_python)) {
  stop("Missing .venv-gdpx4/bin/python. Create it first.")
}
reticulate::use_python(venv_python, required = TRUE)

if (Sys.getenv("HF_TOKEN") == "" && Sys.getenv("HUGGING_FACE_HUB_TOKEN") != "") {
  Sys.setenv(HF_TOKEN = Sys.getenv("HUGGING_FACE_HUB_TOKEN"))
}
if (Sys.getenv("HF_TOKEN") == "" && file.exists(file.path(project_root, ".env"))) {
  env_lines <- readLines(file.path(project_root, ".env"), warn = FALSE)
  token_line <- grep("^(HF_TOKEN|HUGGING_FACE_HUB_TOKEN)=", env_lines, value = TRUE)
  if (length(token_line) > 0) {
    token <- sub("^(HF_TOKEN|HUGGING_FACE_HUB_TOKEN)=", "", token_line[1])
    token <- gsub("^['\"]|['\"]$", "", token)
    Sys.setenv(HF_TOKEN = token)
  }
}
if (Sys.getenv("HF_TOKEN") == "") {
  stop("Token not set. Add HF_TOKEN (or HUGGING_FACE_HUB_TOKEN) to .env or env.")
}

out_dir <- file.path(project_root, "gdpx4-drugseq", "outputs")
fig_dir <- file.path(out_dir, "figures")
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)
```

## Full Lazy Polars Aggregation

```{python full_analysis}
import os
import numpy as np
import pandas as pd
import polars as pl
from huggingface_hub import hf_hub_download

repo_id = "ginkgo-datapoints/GDPx4"
filename = "differential_expression.parquet"
max_rows = int(r.params["max_rows"])
top_n = int(r.params["top_n"])
top_compounds_heatmap = int(r.params["top_compounds_heatmap"])
top_genes_heatmap = int(r.params["top_genes_heatmap"])
target_compound_param = str(r.params["target_compound"] or "").strip()
out_dir = str(r.out_dir)

token = os.getenv("HF_TOKEN") or os.getenv("HUGGING_FACE_HUB_TOKEN")
if not token:
    raise RuntimeError("HF token not found")

local_parquet = hf_hub_download(
    repo_id=repo_id,
    repo_type="dataset",
    filename=filename,
    token=token,
)

lf = pl.scan_parquet(local_parquet)
columns = lf.collect_schema().names()


def pick(cols, candidates):
    m = {c.lower(): c for c in cols}
    for c in candidates:
        if c.lower() in m:
            return m[c.lower()]
    return None

compound_col = pick(columns, ["compound", "perturbation", "drug"])
cell_line_col = pick(columns, ["cell_line", "cellline", "cell_type"])
gene_col = pick(columns, ["gene_symbol", "gene", "feature"])
lfc_col = pick(columns, ["log2FoldChange", "log2_fold_change", "log2fc", "lfc"])
padj_col = pick(columns, ["padj", "adj_pvalue", "qvalue", "fdr"])
container_col = pick(columns, ["container_id", "container", "well"])
concentration_col = pick(columns, ["concentration", "dose", "dose_uM"])

required = [compound_col, gene_col, lfc_col]
if any(c is None for c in required):
    raise RuntimeError(f"Missing required columns in {columns}")

lf_work = lf.head(max_rows) if max_rows and max_rows > 0 else lf

if padj_col:
    sig_expr = ((pl.col(lfc_col).abs() >= 1.0) & (pl.col(padj_col) < 0.05)).cast(pl.Int64)
    min_padj_expr = pl.col(padj_col).min().alias("min_padj")
else:
    sig_expr = pl.lit(0).cast(pl.Int64)
    min_padj_expr = pl.lit(None).alias("min_padj")

compound_aggs = [
    pl.len().alias("rows"),
    pl.col(lfc_col).mean().alias("mean_log2fc"),
    sig_expr.sum().alias("significant_rows"),
]
if container_col:
    compound_aggs.append(pl.col(container_col).n_unique().alias("n_containers"))
if concentration_col:
    compound_aggs.append(pl.col(concentration_col).n_unique().alias("n_concentrations"))

compound_df = (
    lf_work
    .group_by(compound_col)
    .agg(compound_aggs)
    .with_columns((pl.col("significant_rows") / pl.col("rows")).alias("significant_rate"))
    .with_columns(
        pl.when(pl.col("n_containers").is_not_null() & (pl.col("n_containers") > 0))
        .then(pl.col("significant_rows") / pl.col("n_containers"))
        .otherwise(None)
        .alias("significant_per_container")
    )
    .rename({compound_col: "compound"})
    .sort("significant_rate", descending=True)
    .collect()
    .to_pandas()
)

# Gene-by-compound aggregate table for biomarker mining.
gene_compound_aggs = [
    pl.len().alias("rows"),
    pl.col(lfc_col).mean().alias("mean_lfc"),
    pl.col(lfc_col).abs().max().alias("max_abs_lfc"),
    sig_expr.sum().alias("sig_rows"),
    min_padj_expr,
]
if container_col:
    gene_compound_aggs.append(pl.col(container_col).n_unique().alias("n_containers"))
if concentration_col:
    gene_compound_aggs.append(pl.col(concentration_col).n_unique().alias("n_concentrations"))

gene_compound_df = (
    lf_work
    .group_by([gene_col, compound_col])
    .agg(gene_compound_aggs)
    .with_columns((pl.col("sig_rows") / pl.col("rows")).alias("sig_rate"))
    .rename({gene_col: "gene", compound_col: "compound"})
    .collect()
)

# Global recurrence biomarkers: genes repeatedly significant across compounds.
gene_recurrence_df = (
    gene_compound_df.lazy()
    .with_columns((pl.col("sig_rows") > 0).cast(pl.Int64).alias("is_sig_compound"))
    .group_by("gene")
    .agg([
        pl.col("is_sig_compound").sum().alias("compounds_with_signal"),
        pl.col("max_abs_lfc").max().alias("max_abs_lfc"),
        pl.col("max_abs_lfc").mean().alias("mean_abs_lfc"),
    ])
    .with_columns((pl.col("compounds_with_signal") * pl.col("max_abs_lfc")).alias("recurrence_score"))
    .sort("recurrence_score", descending=True)
    .collect()
    .to_pandas()
)

# Pick top active compound and find its biomarker-like genes.
if len(compound_df) == 0:
    raise RuntimeError("No compounds found")

if target_compound_param and (target_compound_param in set(compound_df["compound"].tolist())):
    top_compound = target_compound_param
else:
    top_compound = str(compound_df.iloc[0]["compound"])

dmso_names = [c for c in compound_df["compound"].tolist() if "dmso" in str(c).lower()]
has_dmso = len(dmso_names) > 0

if has_dmso:
    dmso_df = (
        lf_work
        .filter(pl.col(compound_col).cast(pl.Utf8).str.to_lowercase().str.contains("dmso"))
        .group_by(gene_col)
        .agg([
            pl.col(lfc_col).mean().alias("dmso_mean_lfc"),
            pl.len().alias("dmso_rows"),
            *([pl.col(container_col).n_unique().alias("dmso_n_containers")] if container_col else []),
            *([pl.col(concentration_col).n_unique().alias("dmso_n_concentrations")] if concentration_col else []),
        ])
        .rename({gene_col: "gene"})
        .collect()
    )
else:
    dmso_df = pl.DataFrame(
        {
            "gene": pl.Series(name="gene", values=[], dtype=pl.Utf8),
            "dmso_mean_lfc": pl.Series(name="dmso_mean_lfc", values=[], dtype=pl.Float64),
            "dmso_rows": pl.Series(name="dmso_rows", values=[], dtype=pl.Int64),
        }
    )

# Compound-vs-rest specificity model for biomarkers.
target_df = (
    gene_compound_df.lazy()
    .filter(pl.col("compound") == top_compound)
    .select([
        pl.col("gene"),
        pl.col("mean_lfc").alias("target_mean_lfc"),
        pl.col("max_abs_lfc").alias("target_max_abs_lfc"),
        pl.col("sig_rows").alias("target_sig_rows"),
        pl.col("rows").alias("target_rows"),
        pl.col("sig_rate").alias("target_sig_rate"),
        pl.col("min_padj").alias("target_min_padj"),
        *([pl.col("n_containers").alias("target_n_containers")] if container_col else []),
        *([pl.col("n_concentrations").alias("target_n_concentrations")] if concentration_col else []),
    ])
)

other_df = (
    gene_compound_df.lazy()
    .filter(pl.col("compound") != top_compound)
    .group_by("gene")
    .agg([
        pl.col("mean_lfc").abs().mean().alias("other_mean_abs_lfc"),
        pl.col("max_abs_lfc").max().alias("other_max_abs_lfc"),
        (pl.col("sig_rows") > 0).cast(pl.Int64).sum().alias("other_sig_compounds"),
        pl.col("sig_rate").mean().alias("other_mean_sig_rate"),
    ])
)

biomarker_df = (
    target_df
    .join(other_df, on="gene", how="left")
    .join(dmso_df.lazy(), on="gene", how="left")
    .with_columns([
        pl.col("target_mean_lfc").abs().alias("abs_mean_lfc"),
        (pl.col("target_mean_lfc").abs() / (pl.col("other_mean_abs_lfc").fill_null(0.0) + 1e-6)).alias("specificity_ratio"),
        pl.when(pl.col("dmso_mean_lfc").is_not_null())
          .then(pl.col("target_mean_lfc") - pl.col("dmso_mean_lfc"))
          .otherwise(pl.col("target_mean_lfc"))
          .alias("delta_lfc_vs_dmso"),
    ])
    .with_columns([
        pl.when(pl.col("target_min_padj").is_not_null())
          .then((-1.0) * (pl.col("target_min_padj") + 1e-300).log10())
          .otherwise(pl.lit(1.0))
          .alias("target_neg_log10_min_padj"),
    ])
    .with_columns([
        (
            pl.col("abs_mean_lfc")
            * (1 + pl.col("target_sig_rate"))
            * pl.col("target_neg_log10_min_padj")
            * (1 + (pl.col("specificity_ratio") + 1.0).log())
            / (1 + pl.col("other_sig_compounds").fill_null(0))
        ).alias("biomarker_score")
    ])
    .filter(pl.col("target_sig_rows") > 0)
    .filter(pl.col("delta_lfc_vs_dmso").abs() >= 1.0)
    .filter(pl.col("specificity_ratio") >= 1.5)
    .filter(pl.col("other_sig_compounds").fill_null(0) <= 25)
    .sort("biomarker_score", descending=True)
    .head(top_n)
    .collect()
    .to_pandas()
)

# Coexpression: use highest DE biomarker gene as anchor.
if len(biomarker_df) > 0:
    anchor_gene = str(
        biomarker_df.sort_values("abs_mean_lfc", ascending=False).iloc[0]["gene"]
    )
else:
    anchor_gene = str(gene_recurrence_df.iloc[0]["gene"])

sample_cols = [compound_col]
if container_col:
    sample_cols.append(container_col)
if concentration_col:
    sample_cols.append(concentration_col)

gene_sample_lf = (
    lf_work
    .group_by(sample_cols + [gene_col])
    .agg(pl.col(lfc_col).mean().alias("gene_lfc"))
)

anchor_sample_lf = (
    gene_sample_lf
    .filter(pl.col(gene_col) == anchor_gene)
    .select(sample_cols + [pl.col("gene_lfc").alias("anchor_lfc")])
)

coexpr_df = (
    gene_sample_lf
    .join(anchor_sample_lf, on=sample_cols, how="inner")
    .filter(pl.col(gene_col) != anchor_gene)
    .group_by(gene_col)
    .agg([
        pl.corr("gene_lfc", "anchor_lfc").alias("corr"),
        pl.len().alias("n_samples"),
        pl.col("gene_lfc").mean().alias("mean_gene_lfc"),
        pl.col("anchor_lfc").mean().alias("mean_anchor_lfc"),
    ])
    .rename({gene_col: "gene"})
    .with_columns(pl.col("corr").abs().alias("abs_corr"))
    .filter(pl.col("corr").is_not_null())
    .sort("abs_corr", descending=True)
    .collect()
    .to_pandas()
)

top_corr_pairs = coexpr_df.head(40).copy()

# Scatter data for best positive and negative correlated genes.
if len(coexpr_df) > 0:
    pos_gene = coexpr_df.sort_values("corr", ascending=False).iloc[0]["gene"]
    neg_gene = coexpr_df.sort_values("corr", ascending=True).iloc[0]["gene"]
    scatter_genes = list(dict.fromkeys([pos_gene, neg_gene]))
else:
    scatter_genes = []

if scatter_genes:
    scatter_df = (
        gene_sample_lf
        .filter(pl.col(gene_col).is_in(scatter_genes))
        .join(anchor_sample_lf, on=sample_cols, how="inner")
        .select([
            pl.col(gene_col).alias("gene"),
            pl.col("gene_lfc"),
            pl.col("anchor_lfc"),
            *[pl.col(c).cast(pl.Utf8).alias(c) for c in sample_cols],
        ])
        .collect()
        .to_pandas()
    )
else:
    scatter_df = pd.DataFrame(columns=["gene", "gene_lfc", "anchor_lfc"] + sample_cols)

# Heatmap matrix for top compounds x top recurrent genes.
heatmap_compounds = compound_df.head(top_compounds_heatmap)["compound"].tolist()
heatmap_genes = gene_recurrence_df.head(top_genes_heatmap)["gene"].tolist()

heatmap_df = (
    gene_compound_df.lazy()
    .filter(pl.col("compound").is_in(heatmap_compounds) & pl.col("gene").is_in(heatmap_genes))
    .select(["gene", "compound", "mean_lfc", "sig_rate"])
    .collect()
    .to_pandas()
)

rows_processed = int(lf_work.select(pl.len().alias("n")).collect()["n"][0])
unique_compounds = int(lf_work.select(pl.col(compound_col).n_unique().alias("n")).collect()["n"][0])
unique_genes = int(lf_work.select(pl.col(gene_col).n_unique().alias("n")).collect()["n"][0])
if cell_line_col:
    unique_cell_lines = int(lf_work.select(pl.col(cell_line_col).n_unique().alias("n")).collect()["n"][0])
    cell_lines = lf_work.select(pl.col(cell_line_col).unique().sort()).collect().to_series(0).to_list()
else:
    unique_cell_lines = 0
    cell_lines = []

summary_df = pd.DataFrame([
    {
        "rows_processed": rows_processed,
        "unique_compounds": unique_compounds,
        "unique_cell_lines": unique_cell_lines,
        "unique_genes": unique_genes,
        "cell_lines": ", ".join(cell_lines),
        "repo_id": repo_id,
        "file": filename,
        "target_compound": top_compound,
        "anchor_gene_for_coexpression": anchor_gene,
        "n_dmso_compounds_detected": len(dmso_names),
    }
])

compound_df.to_csv(os.path.join(out_dir, "gdpx4_compound_summary.csv"), index=False)
gene_recurrence_df.to_csv(os.path.join(out_dir, "gdpx4_gene_recurrence.csv"), index=False)
biomarker_df.to_csv(os.path.join(out_dir, "gdpx4_candidate_biomarkers.csv"), index=False)
coexpr_df.to_csv(os.path.join(out_dir, "gdpx4_anchor_gene_coexpression.csv"), index=False)
top_corr_pairs.to_csv(os.path.join(out_dir, "gdpx4_anchor_gene_coexpression_top40.csv"), index=False)
scatter_df.to_csv(os.path.join(out_dir, "gdpx4_anchor_gene_scatter.csv"), index=False)
heatmap_df.to_csv(os.path.join(out_dir, "gdpx4_heatmap_long.csv"), index=False)
summary_df.to_csv(os.path.join(out_dir, "gdpx4_rmd_run_summary.csv"), index=False)
```

## Run Summary

```{r}
summary_tbl <- read.csv(file.path(out_dir, "gdpx4_rmd_run_summary.csv"), stringsAsFactors = FALSE)
summary_tbl
```

## Plot 1: Significant DE Genes Per Compound

```{r}
compound_tbl <- read.csv(file.path(out_dir, "gdpx4_compound_summary.csv"), stringsAsFactors = FALSE)
compound_tbl <- compound_tbl[order(-compound_tbl$significant_rows), ]
compound_tbl$compound <- reorder(compound_tbl$compound, compound_tbl$significant_rows)

p1 <- ggplot(compound_tbl, aes(x = compound, y = significant_rows)) +
  geom_col(fill = "#2E6F95") +
  coord_flip() +
  labs(
    title = "Significant DE Genes by Compound",
    subtitle = "Threshold: |log2FC| >= 1 and padj < 0.05",
    x = NULL,
    y = "Significant genes"
  ) +
  theme_minimal(base_size = 12)

p1

ggsave(file.path(fig_dir, "gdpx4_significant_genes_by_compound.png"), p1, width = 10, height = 6, dpi = 150)
```

## Plot 2: Significant Rate by Compound

```{r}
p2 <- ggplot(compound_tbl, aes(x = reorder(compound, significant_rate), y = significant_rate)) +
  geom_col(fill = "#7AAE42") +
  coord_flip() +
  scale_y_continuous(labels = function(x) paste0(round(100 * x, 1), "%")) +
  labs(
    title = "Significant DE Rate by Compound",
    x = NULL,
    y = "Significant rate"
  ) +
  theme_minimal(base_size = 12)

p2

ggsave(file.path(fig_dir, "gdpx4_significant_rate_by_compound.png"), p2, width = 10, height = 6, dpi = 150)
```

## Plot 3: Cross-Compound Gene Recurrence (Top 25)

```{r}
rec_tbl <- read.csv(file.path(out_dir, "gdpx4_gene_recurrence.csv"), stringsAsFactors = FALSE)
rec25 <- rec_tbl[1:min(25, nrow(rec_tbl)), ]
rec25$gene <- factor(rec25$gene, levels = rev(rec25$gene))

p3 <- ggplot(rec25, aes(x = gene, y = compounds_with_signal, fill = recurrence_score)) +
  geom_col() +
  coord_flip() +
  scale_fill_gradient(low = "#a6cee3", high = "#1f78b4") +
  labs(
    title = "Top Recurrent Response Genes Across Compounds",
    x = NULL,
    y = "Compounds with signal",
    fill = "Recurrence score"
  ) +
  theme_minimal(base_size = 11)

p3

ggsave(file.path(fig_dir, "gdpx4_gene_recurrence_top25.png"), p3, width = 10, height = 7, dpi = 150)
```

## Plot 4: Candidate Biomarkers for Top Active Compound

```{r}
bio_tbl <- read.csv(file.path(out_dir, "gdpx4_candidate_biomarkers.csv"), stringsAsFactors = FALSE)
if (nrow(bio_tbl) > 0) {
  top_label <- summary_tbl$target_compound[1]
  bio_show <- bio_tbl[1:min(20, nrow(bio_tbl)), ]
  bio_show$gene <- factor(bio_show$gene, levels = rev(bio_show$gene))

  p4 <- ggplot(bio_show, aes(x = gene, y = biomarker_score, fill = abs_mean_lfc)) +
    geom_col() +
    coord_flip() +
    scale_fill_gradient(low = "#fee8c8", high = "#e34a33") +
    labs(
      title = paste0("Candidate Biomarkers for ", top_label),
      subtitle = "Score uses target-vs-DMSO delta, confidence, consistency, and compound-vs-rest specificity",
      x = NULL,
      y = "Biomarker score",
      fill = "|Mean LFC|"
    ) +
    theme_minimal(base_size = 11)

  print(p4)
  ggsave(file.path(fig_dir, "gdpx4_candidate_biomarkers_top20.png"), p4, width = 10, height = 7, dpi = 150)
} else {
  cat("No biomarker candidates found under current thresholds.")
}
```

## Plot 5: Target-vs-DMSO Delta vs Specificity

```{r}
if (nrow(bio_tbl) > 0 && "delta_lfc_vs_dmso" %in% colnames(bio_tbl)) {
  scatter_tbl <- bio_tbl[1:min(40, nrow(bio_tbl)), ]
  scatter_tbl$label <- as.character(scatter_tbl$gene)

  p5 <- ggplot(scatter_tbl, aes(x = specificity_ratio, y = abs(delta_lfc_vs_dmso))) +
    geom_point(aes(size = biomarker_score, color = target_neg_log10_min_padj), alpha = 0.85) +
    geom_text(aes(label = label), hjust = 0, nudge_x = 0.05, size = 3, check_overlap = TRUE) +
    labs(
      title = paste0("Biomarker Landscape for ", summary_tbl$target_compound[1]),
      subtitle = "Top candidates: specificity vs |target - DMSO| effect",
      x = "Specificity ratio (target vs rest)",
      y = "|Delta log2FC vs DMSO|",
      color = "-log10(min padj)",
      size = "Biomarker score"
    ) +
    theme_minimal(base_size = 11)

  p5
  ggsave(file.path(fig_dir, "gdpx4_biomarker_delta_vs_specificity.png"), p5, width = 11, height = 8, dpi = 150)
}
```

## Plot 6: Heatmap (Top Recurrent Genes x Top Active Compounds)

```{r}
heat_tbl <- read.csv(file.path(out_dir, "gdpx4_heatmap_long.csv"), stringsAsFactors = FALSE)
if (nrow(heat_tbl) > 0) {
  # Order genes by average absolute effect in this displayed matrix
  gene_order <- aggregate(abs(mean_lfc) ~ gene, data = heat_tbl, FUN = mean)
  gene_order <- gene_order[order(gene_order[,2], decreasing = TRUE), "gene"]
  comp_order <- aggregate(sig_rate ~ compound, data = heat_tbl, FUN = mean)
  comp_order <- comp_order[order(comp_order[,2], decreasing = TRUE), "compound"]

  heat_tbl$gene <- factor(heat_tbl$gene, levels = gene_order)
  heat_tbl$compound <- factor(heat_tbl$compound, levels = comp_order)

  p5 <- ggplot(heat_tbl, aes(x = compound, y = gene, fill = mean_lfc)) +
    geom_tile() +
    scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0) +
    labs(
      title = "Mean log2FC Heatmap: Top Recurrent Genes x Top Compounds",
      x = "Compound",
      y = "Gene",
      fill = "Mean log2FC"
    ) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

  p5
  ggsave(file.path(fig_dir, "gdpx4_heatmap_top_genes_compounds.png"), p5, width = 11, height = 9, dpi = 150)
} else {
  cat("No heatmap rows available.")
}
```

## Plot 7: Coexpression With Anchor Gene (Top |r| Genes)

```{r}
co_tbl <- read.csv(file.path(out_dir, "gdpx4_anchor_gene_coexpression_top40.csv"), stringsAsFactors = FALSE)
if (nrow(co_tbl) > 0) {
  co_show <- co_tbl[1:min(30, nrow(co_tbl)), ]
  co_show$gene <- factor(co_show$gene, levels = rev(co_show$gene))

  p7 <- ggplot(co_show, aes(x = gene, y = corr, fill = corr)) +
    geom_col() +
    coord_flip() +
    scale_fill_gradient2(low = "#2166ac", mid = "#f7f7f7", high = "#b2182b", midpoint = 0) +
    labs(
      title = paste0("Top Coexpressed Genes With ", summary_tbl$anchor_gene_for_coexpression[1]),
      subtitle = "Correlation across compound-container-concentration sample contexts",
      x = NULL,
      y = "Correlation (r)",
      fill = "r"
    ) +
    theme_minimal(base_size = 11)

  print(p7)
  ggsave(file.path(fig_dir, "gdpx4_anchor_gene_top_coexpression.png"), p7, width = 10, height = 8, dpi = 150)
}
```

## Plot 8: Anchor-Gene Scatter (Best Positive/Negative Correlates)

```{r}
scatter_long <- read.csv(file.path(out_dir, "gdpx4_anchor_gene_scatter.csv"), stringsAsFactors = FALSE)
if (nrow(scatter_long) > 0) {
  p8 <- ggplot(scatter_long, aes(x = anchor_lfc, y = gene_lfc)) +
    geom_point(alpha = 0.35) +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.7) +
    facet_wrap(~gene, scales = "free_y") +
    labs(
      title = paste0("Anchor-Gene Correlation Scatter: ", summary_tbl$anchor_gene_for_coexpression[1]),
      x = "Anchor gene log2FC",
      y = "Gene log2FC"
    ) +
    theme_minimal(base_size = 11)

  print(p8)
  ggsave(file.path(fig_dir, "gdpx4_anchor_gene_scatter_facets.png"), p8, width = 11, height = 5, dpi = 150)
}
```
