---
title: "PBMC 3k scRNA-seq Pipeline (Arrow + Seurat)"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
# Install dependencies via pak (preferred)
# Set SKIP_PAK_INSTALL=1 to skip install during render.
# Set SKIP_R_RUN=1 to render without executing code chunks.
required_pkgs <- c("Seurat", "Matrix", "ggplot2", "dplyr", "tidyr", "arrow", "DESeq2")
if (!requireNamespace("pak", quietly = TRUE)) {
  install.packages("pak")
}
if (Sys.getenv("SKIP_PAK_INSTALL") != "1") {
  missing_pkgs <- required_pkgs[!sapply(required_pkgs, requireNamespace, quietly = TRUE)]
  if (length(missing_pkgs) > 0) {
    pak::pkg_install(missing_pkgs)
  }
}

if (Sys.getenv("SKIP_R_RUN") == "1") {
  knitr::opts_chunk$set(eval = FALSE)
} else {
  library(Seurat)
  library(Matrix)
  library(ggplot2)
  library(dplyr)
  library(arrow)
  library(tidyr)
}

DATASET <- "mannekuntanagendra/single-cell-rna-seq-analysis-qc-clustering-pbmc-3k"
DATA_DIR <- "data/pbmc3k"
OUTPUT_DIR <- "outputs"
ARROW_DIR <- file.path(OUTPUT_DIR, "arrow")

if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR, recursive = TRUE)
if (!dir.exists(ARROW_DIR)) dir.create(ARROW_DIR, recursive = TRUE)
```

## Summary

End‑to‑end scRNA‑seq workflow on PBMC 3k: QC, normalization, feature selection, PCA/UMAP, clustering, marker discovery, and export of reusable artifacts.

## Get the Data

If you have Kaggle CLI configured, this will download and unzip the dataset. Otherwise, manually download from Kaggle and place the files inside `data/pbmc3k/`.

**Why this matters:** Data access should be reproducible and boring. This shows an automated download path with a manual fallback.

```{r kaggle, message=FALSE, warning=FALSE}
ensure_data_dir <- function(path) {
  if (!dir.exists(path)) dir.create(path, recursive = TRUE)
}

has_kaggle_token <- function() {
  file.exists(file.path(path.expand("~"), ".kaggle", "kaggle.json"))
}

try_kaggle_download <- function(dataset, dest_dir) {
  ensure_data_dir(dest_dir)

  kaggle_bin <- Sys.which("kaggle")
  if (kaggle_bin == "" || !has_kaggle_token()) {
    return(invisible(FALSE))
  }

  args <- c("datasets", "download", "-d", dataset, "-p", dest_dir, "--unzip")
  system2(kaggle_bin, args = args)
  invisible(TRUE)
}

try_kaggle_download(DATASET, DATA_DIR)

list.files(DATA_DIR, recursive = TRUE)[1:10]
```

## Load Dataset

This attempts to load a 10x directory if present. The Kaggle PBMC 3k dataset typically ships with `matrix.mtx`, `features.tsv`, and `barcodes.tsv`.

**Why this matters:** 10x Genomics is the most common scRNA‑seq format, so it’s worth handling cleanly.

```{r load}
load_dataset_auto <- function(data_dir) {
  mtx <- list.files(data_dir, pattern = "matrix.mtx", recursive = TRUE, full.names = TRUE)
  if (length(mtx) > 0) {
    return(Read10X(data.dir = dirname(mtx[1])))
  }

  stop("No supported files found. Expected 10x mtx directory.")
}

counts <- load_dataset_auto(DATA_DIR)
seu <- CreateSeuratObject(counts = counts, project = "pbmc3k")

seu
```

## Basic QC

**What happens here:** Compute per‑cell QC metrics and look at the distributions (library size, gene counts, mitochondrial fraction).

**Why this matters:** QC is where you separate biology from technical artifacts.

**Notes:** Mitochondrial percentage, total UMIs, and detected genes are standard diagnostics for cell viability and doublets. These plots make it easier to choose sensible thresholds.

```{r qc}
seu[["percent.mt"]] <- PercentageFeatureSet(seu, pattern = "^MT-")

VlnPlot(seu, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
FeatureScatter(seu, feature1 = "nCount_RNA", feature2 = "percent.mt")

qc_tbl <- seu@meta.data %>%
  select(nFeature_RNA, nCount_RNA, percent.mt)
qc_tbl$cell_id <- rownames(seu@meta.data)

arrow::write_feather(qc_tbl, file.path(ARROW_DIR, "qc_metrics.feather"))
```

## Filter, Normalize, and Find Variable Features

**What happens here:** Filter low‑quality cells/genes, normalize counts, and select highly variable genes for downstream signal.

**Why this matters:** These are the standard preprocessing choices for scRNA‑seq.

**Notes:** Filtering removes sparse or damaged cells; log‑normalization stabilizes variance; highly variable genes capture the most informative biology and improve downstream dimensionality reduction.

```{r normalize}
seu <- subset(seu, subset = nFeature_RNA > 200 & nFeature_RNA < 2500 & percent.mt < 15)

seu <- NormalizeData(seu)
seu <- FindVariableFeatures(seu, selection.method = "vst", nfeatures = 2000)
seu <- ScaleData(seu)
```

## PCA, UMAP, Clustering

**What happens here:** Reduce dimensionality (PCA), build a neighborhood graph, embed with UMAP, and identify clusters.

**Why this matters:** Dimensionality reduction + clustering are the backbone of scRNA‑seq analysis.

**Notes:** PCA captures major axes of variation; a KNN graph models local structure; UMAP provides a 2D view; Leiden clustering detects communities that often map to cell types or states.

```{r reduce}
seu <- RunPCA(seu)
seu <- FindNeighbors(seu, dims = 1:20)
seu <- FindClusters(seu, resolution = 0.6)
seu <- RunUMAP(seu, dims = 1:20)

DimPlot(seu, reduction = "umap", label = TRUE)
```

## Marker Genes

**What happens here:** Identify genes that are differentially expressed in each cluster.

**Why this matters:** Marker discovery drives cell‑type labeling and interpretation.

**Notes:** Differential expression across clusters highlights signature genes that can be matched to known cell types or used to discover novel states.

```{r markers}
message("Running DESeq2 marker tests (output condensed)...")
invisible(capture.output(
  markers <- suppressMessages(FindAllMarkers(
    seu,
    only.pos = TRUE,
    min.pct = 0.25,
    logfc.threshold = 0.25,
    test.use = "DESeq2",
    verbose = FALSE
  )),
  file = nullfile()
))
head(markers, 10)
```

## DESeq2 Marker Plot

**What happens here:** Visualize DESeq2 marker test significance per cluster.

**Why this matters:** It makes the DESeq2-based marker ranking visible rather than only showing raw tables.

```{r deseq2_marker_plot, fig.width=10, fig.height=7}
marker_plot_df <- markers %>%
  filter(!is.na(p_val_adj), !is.na(avg_log2FC)) %>%
  mutate(
    cluster = factor(cluster),
    neg_log10_padj = -log10(pmax(p_val_adj, 1e-300))
  )

label_df <- marker_plot_df %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 2, with_ties = FALSE) %>%
  ungroup()

ggplot(marker_plot_df, aes(x = avg_log2FC, y = neg_log10_padj)) +
  geom_point(alpha = 0.35, size = 0.7, color = "#1f78b4") +
  facet_wrap(~ cluster, scales = "free_y") +
  geom_text(
    data = label_df,
    aes(label = gene),
    size = 2.4,
    vjust = -0.5,
    check_overlap = TRUE
  ) +
  labs(
    title = "DESeq2 marker significance by cluster",
    x = "avg_log2FC",
    y = "-log10(adjusted p-value)"
  ) +
  theme_bw()
```

## Putative Cell‑Type Labels (Manual)

**What happens here:** Use canonical PBMC marker sets and cluster‑level scores to assign **putative** labels.  
These are best‑effort labels based on marker expression, not ground truth.

```{r cell_type_labels}
marker_sets <- list(
  B_cells = c("MS4A1", "CD79A", "CD74"),
  T_cells = c("CD3D", "CD3E", "IL7R", "LTB"),
  NK_cells = c("NKG7", "GNLY", "PRF1"),
  Monocytes = c("LST1", "S100A8", "S100A9", "LYZ"),
  Dendritic = c("FCER1A", "CST3"),
  Platelets = c("PPBP", "PF4")
)

# Score each marker set per cell
seu <- AddModuleScore(seu, features = marker_sets, name = "celltype_score")

# Map scores back to names
score_cols <- paste0("celltype_score", seq_along(marker_sets))
score_df <- FetchData(seu, vars = c("seurat_clusters", score_cols))
colnames(score_df) <- c("cluster", names(marker_sets))

# Average score per cluster and assign the max
cluster_scores <- score_df %>%
  group_by(cluster) %>%
  summarise(across(all_of(names(marker_sets)), mean), .groups = "drop")

cluster_labels <- cluster_scores %>%
  pivot_longer(-cluster, names_to = "label", values_to = "score") %>%
  group_by(cluster) %>%
  slice_max(score, n = 1, with_ties = FALSE) %>%
  ungroup()

cluster_labels
```

```{r apply_labels}
label_map <- setNames(cluster_labels$label, cluster_labels$cluster)
cell_type_vec <- label_map[as.character(seu$seurat_clusters)]
names(cell_type_vec) <- colnames(seu)
seu <- AddMetaData(seu, metadata = cell_type_vec, col.name = "cell_type")
DimPlot(seu, reduction = "umap", group.by = "cell_type", label = TRUE)
```

## Selected Gene Boxplots

**What happens here:** Visualize expression of canonical immune markers across clusters.

**Why this matters:** These plots help validate cluster identities.

**Why these genes:** `MS4A1` (B cells), `CD3D` (T cells), `LST1` (monocytes), and `NKG7`/`GNLY` (NK/cytotoxic cells) are canonical PBMC markers that usually separate major immune lineages.

**Notes:** Boxplots summarize expression distributions and outliers, making cluster‑level differences easy to compare.

```{r gene_boxplots}
marker_genes <- c("MS4A1", "CD3D", "LST1", "NKG7", "GNLY")

gene_df <- FetchData(seu, vars = c("seurat_clusters", marker_genes))
gene_long <- gene_df %>%
  pivot_longer(cols = all_of(marker_genes), names_to = "gene", values_to = "expression")

ggplot(gene_long, aes(x = seurat_clusters, y = expression, fill = seurat_clusters)) +
  geom_boxplot(outlier.size = 0.4, width = 0.6, alpha = 0.85) +
  facet_wrap(~ gene, scales = "free_y", ncol = 3) +
  labs(x = "Cluster", y = "Expression") +
  scale_fill_brewer(palette = "Set3") +
  theme_bw() +
  theme(legend.position = "none")
```

## Save Outputs

**What happens here:** Save a Seurat object for reuse and export lightweight Arrow tables for downstream consumption.

**Why this matters:** Keeps the results portable, not just in plots.

```{r save}
saveRDS(seu, file.path(OUTPUT_DIR, "pbmc3k_seurat.rds"))

cluster_tbl <- seu@meta.data %>%
  select(nFeature_RNA, nCount_RNA, percent.mt, seurat_clusters)
cluster_tbl$cell_id <- rownames(seu@meta.data)

arrow::write_feather(cluster_tbl, file.path(ARROW_DIR, "cell_clusters.feather"))
```
