---
title: "Breast Cancer Cell Lines Atlas — Single-Cell RNA-seq"
output:
  html_document:
    toc: true
    toc_depth: 2
    theme: readable
    highlight: tango
---

```{r setup, message=FALSE, warning=FALSE}
set.seed(7)
options(stringsAsFactors = FALSE)

RUN_MARKERS <- TRUE

suppressPackageStartupMessages({
  library(Seurat)
  library(Matrix)
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(tidyr)
  library(stringr)
  library(biomaRt)
  library(viridis)
})

if (!requireNamespace("xgboost", quietly = TRUE)) {
  stop("This notebook requires xgboost for the ML section. Install with pak::pkg_install('xgboost').")
}

root_dir <- normalizePath(file.path(".."))
mtx_dir <- file.path(root_dir, "data", "MTX_FMT_GSE173634")
mtx_file <- file.path(mtx_dir, "matrix.mtx")
features_file <- file.path(mtx_dir, "features.tsv")
barcodes_file <- file.path(mtx_dir, "barcodes.tsv")

if (!file.exists(mtx_file)) {
  stop("Missing MTX files. Download the Kaggle dataset into breast-cancer-scrna/data.")
}
```

# Why this dataset

This is a single-cell atlas of 32 human breast cancer cell lines (35,276 cells) originally published as GSE173634. I picked it because it’s a clean, well-annotated dataset with a wide range of molecular subtypes. That makes it a good playground for testing clustering quality, state scoring, and marker-driven interpretation.

# Glossary (abbreviations)

- `scRNA-seq`: single-cell RNA sequencing.
- `QC`: quality control.
- `UMI`: unique molecular identifier, used to count transcripts.
- `nCount_RNA`: total UMI counts per cell.
- `nFeature_RNA`: number of genes detected per cell.
- `percent.mt`: percent of counts from mitochondrial genes per cell.
- `HVG`: highly variable genes.
- `PCA`: principal component analysis.
- `UMAP`: Uniform Manifold Approximation and Projection (2D embedding).
- `S/G2M`: cell-cycle phases.
- `HVG`: highly variable genes used for dimensionality reduction.
- `Epithelial/Mesenchymal`: broad expression states often contrasted in cancer.
- `Markers`: genes enriched in a cluster relative to others.

# What you can learn from each plot

- `QC violin plots`: distributions of `nFeature_RNA`, `nCount_RNA`, and `percent.mt` to detect low-quality cells or outliers.
- `Variable feature plot`: whether HVGs are captured and how strongly they vary.
- `Elbow plot`: how many PCs capture most signal before diminishing returns.
- `UMAP (clusters)`: overall structure and separation of clusters.
- `UMAP (cell line)`: whether clustering aligns with known cell-line identity.
- `UMAP (cell cycle)`: whether clusters are driven by cycling status.
- `State score feature plots`: spatial pattern of epithelial/mesenchymal/proliferative/stress programs.
- `State-labeled UMAP`: high-level interpretation of clusters by dominant program.
- `Gene boxplots`: per-cluster expression for key genes (signals specific biology).
- `Validation scatter/UMAP/VLN plots`: check if clusters are dominated by QC artifacts.

# What we learn from the full analysis

This pipeline shows how breast cancer cell lines separate into interpretable expression states, which programs dominate each cluster, and whether the structure is biological rather than QC-driven. It also highlights which canonical genes (e.g., `EPCAM`, `VIM`, `MKI67`, `ERBB2`, `ESR1`) anchor those states, giving you a concrete story you can explain to a technical audience.

# Load counts

We start from the Matrix Market files bundled with the dataset. Barcodes embed the cell-line label, so we recover it directly from the IDs.

```{r load-data, message=FALSE, warning=FALSE}
bc_counts <- ReadMtx(mtx = mtx_file, features = features_file, cells = barcodes_file, feature.column = 1)

bc <- CreateSeuratObject(counts = bc_counts, min.cells = 3, min.features = 200)
bc$cell_line <- sub("_.*", "", colnames(bc))

bc
```

```{r inspect-metadata}
head(bc@meta.data)
```

# QC and filtering

I keep the QC story simple: remove empty droplets and clear low-quality cells. I’m not doing aggressive filtering because these are cell lines, but I still want to avoid obvious junk.

```{r qc-metrics, fig.width=11, fig.height=4}
bc[["percent.mt"]] <- PercentageFeatureSet(bc, pattern = "^MT-")

VlnPlot(bc, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

```{r qc-filter}
bc <- subset(bc, subset = nFeature_RNA > 300 & nFeature_RNA < 7000 & percent.mt < 20)

bc
```

# Normalize + feature selection

I use log-normalization and select highly variable genes. It’s fast and gives clean separation for this scale of data.

```{r normalize}
bc <- NormalizeData(bc)
bc <- FindVariableFeatures(bc, selection.method = "vst", nfeatures = 3000)

VariableFeaturePlot(bc)
```

# Dimensionality reduction and clustering

PCA → UMAP → clustering. I keep the resolution moderate to avoid splitting cell lines into tiny micro-clusters unless the data strongly supports it.

```{r pca-umap, fig.width=8, fig.height=5}
bc <- ScaleData(bc, features = VariableFeatures(bc))
bc <- RunPCA(bc, features = VariableFeatures(bc))

ElbowPlot(bc)
```

```{r cluster, fig.width=8, fig.height=6}
bc <- FindNeighbors(bc, dims = 1:30)
bc <- FindClusters(bc, resolution = 0.4)
bc <- RunUMAP(bc, dims = 1:30)

DimPlot(bc, reduction = "umap", label = TRUE)
```

# Cell line context

The barcode prefix encodes cell-line identity, so I keep it as a grouping label.

```{r cell-line-umap, fig.width=9, fig.height=6}
DimPlot(bc, reduction = "umap", group.by = "cell_line", label = FALSE) +
  guides(color = guide_legend(override.aes = list(size = 3), ncol = 2))
```

# Cell-cycle scoring

This adds some biological context that’s often hidden in raw clusters.

```{r biomart-setup}
mart <- NULL
for (mirror in c("useast", "uswest", "asia", "www")) {
  mart <- tryCatch(
    biomaRt::useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl", mirror = mirror),
    error = function(e) NULL
  )
  if (!is.null(mart)) break
}

if (is.null(mart)) {
  stop("Unable to connect to Ensembl via biomaRt. Try again later or set a different mirror.")
}

map_symbols <- function(symbols) {
  symbol_map <- biomaRt::getBM(
    attributes = c("hgnc_symbol", "ensembl_gene_id"),
    filters = "hgnc_symbol",
    values = unique(symbols),
    mart = mart
  )

  symbol_map <- symbol_map %>%
    dplyr::filter(hgnc_symbol != "", ensembl_gene_id %in% rownames(bc)) %>%
    dplyr::distinct(hgnc_symbol, .keep_all = TRUE)

  setNames(symbol_map$ensembl_gene_id, symbol_map$hgnc_symbol)
}
```

```{r cell-cycle, fig.width=8, fig.height=6}
cc <- Seurat::cc.genes

s_genes <- unname(map_symbols(cc$s.genes))
g2m_genes <- unname(map_symbols(cc$g2m.genes))

if (length(s_genes) >= 10 && length(g2m_genes) >= 10) {
  bc <- CellCycleScoring(bc, s.features = s_genes, g2m.features = g2m_genes)
  DimPlot(bc, reduction = "umap", group.by = "Phase")
} else {
  message("Cell-cycle scoring skipped: not enough cell-cycle genes matched.")
}
```

# Gene symbol mapping

The count matrix uses Ensembl IDs. For the state scores and plots below, I map a small set of gene symbols to their Ensembl IDs.

```{r gene-symbol-map}
symbol_sets <- list(
  Epithelial = c("EPCAM", "KRT8", "KRT18", "KRT19", "MUC1"),
  Mesenchymal = c("VIM", "FN1", "ZEB1", "ZEB2", "COL1A1"),
  Proliferative = c("MKI67", "TOP2A", "HMGB2", "TYMS", "STMN1"),
  Stress = c("FOS", "JUN", "HSPA1A", "HSP90AA1"),
  PlotGenes = c("EPCAM", "KRT19", "VIM", "MKI67", "TOP2A", "ERBB2", "ESR1")
)

target_symbols <- unique(unlist(symbol_sets))

symbol_to_ens <- map_symbols(target_symbols)
```

# XGBoost: Cell-line prediction

Train a multiclass XGBoost model to predict cell-line identity from expression profiles.

```{r xgboost-cell-line}
set.seed(7)

ml_features <- head(VariableFeatures(bc), 400)
if (length(ml_features) < 50) {
  stop("Not enough variable features available for XGBoost.")
}

feature_map <- biomaRt::getBM(
  attributes = c("ensembl_gene_id", "hgnc_symbol"),
  filters = "ensembl_gene_id",
  values = unique(ml_features),
  mart = mart
) %>%
  dplyr::filter(hgnc_symbol != "") %>%
  dplyr::distinct(ensembl_gene_id, .keep_all = TRUE)

feature_label_map <- setNames(feature_map$hgnc_symbol, feature_map$ensembl_gene_id)
feature_display <- ifelse(
  ml_features %in% names(feature_label_map),
  feature_label_map[ml_features],
  ml_features
)
feature_display <- make.unique(feature_display)

ml_df <- FetchData(bc, vars = c("cell_line", ml_features))
colnames(ml_df) <- c("cell_line", feature_display)

# Keep runtime practical by balancing and capping cells per line.
ml_df <- ml_df %>%
  mutate(row_id = dplyr::row_number()) %>%
  group_by(cell_line) %>%
  mutate(.rand = runif(dplyr::n())) %>%
  arrange(.rand, .by_group = TRUE) %>%
  mutate(.grp_rank = row_number()) %>%
  filter(.grp_rank <= 150) %>%
  dplyr::select(-.rand, -.grp_rank) %>%
  ungroup()

train_idx <- ml_df %>%
  group_by(cell_line) %>%
  group_map(~ sample(.x$row_id, size = max(1, floor(0.8 * nrow(.x))))) %>%
  unlist(use.names = FALSE)

labels_factor <- factor(ml_df$cell_line)
labels_num <- as.integer(labels_factor) - 1

x_mat <- as.matrix(ml_df[, feature_display, drop = FALSE])
train_mask <- ml_df$row_id %in% train_idx

dtrain <- xgboost::xgb.DMatrix(data = x_mat[train_mask, , drop = FALSE], label = labels_num[train_mask])
dtest <- xgboost::xgb.DMatrix(data = x_mat[!train_mask, , drop = FALSE], label = labels_num[!train_mask])

params <- list(
  objective = "multi:softprob",
  num_class = nlevels(labels_factor),
  eval_metric = "mlogloss",
  eta = 0.15,
  max_depth = 4,
  subsample = 0.8,
  colsample_bytree = 0.8
)

xgb_model <- xgboost::xgb.train(
  params = params,
  data = dtrain,
  nrounds = 60,
  watchlist = list(train = dtrain, test = dtest),
  early_stopping_rounds = 8,
  verbose = 0
)

pred_prob <- matrix(
  predict(xgb_model, dtest),
  ncol = nlevels(labels_factor),
  byrow = TRUE
)
pred_num <- max.col(pred_prob) - 1
truth_num <- labels_num[!train_mask]

xgb_accuracy <- mean(pred_num == truth_num)
xgb_accuracy

pred_lab <- factor(pred_num + 1, levels = seq_len(nlevels(labels_factor)), labels = levels(labels_factor))
truth_lab <- factor(truth_num + 1, levels = seq_len(nlevels(labels_factor)), labels = levels(labels_factor))

conf_mat <- table(truth = truth_lab, pred = pred_lab)
conf_mat
```

# XGBoost: Top predictive genes

Rank genes by XGBoost gain importance to identify strong differentiators between cell lines.

```{r xgboost-importance, fig.width=9, fig.height=6}
imp <- xgboost::xgb.importance(model = xgb_model)
top_imp <- imp %>%
  as_tibble() %>%
  dplyr::slice_head(n = 25)

top_imp %>%
  dplyr::select(Feature, Gain, Cover, Frequency)

ggplot(top_imp, aes(x = reorder(Feature, Gain), y = Gain)) +
  geom_col(fill = "#1f78b4", alpha = 0.9) +
  coord_flip() +
  labs(
    title = "Top XGBoost features for cell-line classification",
    x = "Gene symbol (feature)",
    y = "Gain importance"
  ) +
  theme_bw()
```

# State scoring (epithelial vs mesenchymal)

These scores are not “cell types” in the classical sense, but they help interpret how clusters lean along common cancer state axes.

```{r state-scores, fig.width=10, fig.height=8}
state_sets <- lapply(symbol_sets[names(symbol_sets) != "PlotGenes"], function(g) unname(symbol_to_ens[g]))
state_sets <- lapply(state_sets, function(g) g[!is.na(g)])
state_sets <- state_sets[lengths(state_sets) > 0]

if (length(state_sets) == 0) {
  stop("No state gene sets matched the dataset.")
}

bc <- AddModuleScore(bc, features = state_sets, name = names(state_sets))

score_cols <- paste0(names(state_sets), "1")
score_cols <- score_cols[score_cols %in% colnames(bc@meta.data)]

if (length(score_cols) > 0) {
FeaturePlot(bc, features = score_cols, ncol = 2)
} else {
  message("State scores skipped: no score columns were created.")
}
```

```{r state-labels, fig.width=8, fig.height=6}
if (length(score_cols) > 0) {
  score_df <- bc@meta.data %>%
    dplyr::select(seurat_clusters, all_of(score_cols)) %>%
    dplyr::group_by(seurat_clusters) %>%
    dplyr::summarize(across(all_of(score_cols), mean), .groups = "drop")

  score_mat <- as.data.frame(score_df)
  row.names(score_mat) <- score_mat$seurat_clusters
  score_mat$seurat_clusters <- NULL

  state_labels <- apply(score_mat, 1, function(x) names(state_sets)[which.max(x)])

  bc$state_label <- unname(state_labels[as.character(bc$seurat_clusters)])

  DimPlot(bc, reduction = "umap", group.by = "state_label", label = TRUE)
} else {
  message("State labels skipped: no scores available.")
}
```

# Marker discovery

Markers help explain *why* a cluster exists. This step is optional because it can be slow on laptops. Toggle `RUN_MARKERS` in the setup chunk if you want to run it.

```{r markers}
if (RUN_MARKERS) {
  if (!requireNamespace("DESeq2", quietly = TRUE)) {
    stop("RUN_MARKERS=TRUE requires DESeq2. Install it with pak::pkg_install('DESeq2').")
  }

  bc_markers <- subset(bc, downsample = 8)

  message("Running DESeq2 marker tests (output condensed)...")
  invisible(capture.output(
    markers <- suppressMessages(FindAllMarkers(
      bc_markers,
      only.pos = TRUE,
      min.pct = 0.15,
      logfc.threshold = 0.3,
      test.use = "DESeq2",
      verbose = FALSE
    )),
    file = nullfile()
  ))

  top_markers <- markers %>%
    group_by(cluster) %>%
    slice_max(avg_log2FC, n = 5) %>%
    ungroup()

  marker_map <- biomaRt::getBM(
    attributes = c("ensembl_gene_id", "hgnc_symbol"),
    filters = "ensembl_gene_id",
    values = unique(top_markers$gene),
    mart = mart
  )

  top_markers_tbl <- top_markers %>%
    dplyr::left_join(marker_map, by = c("gene" = "ensembl_gene_id")) %>%
    dplyr::mutate(gene_label = ifelse(hgnc_symbol != "", hgnc_symbol, gene)) %>%
    dplyr::select(cluster, gene_label, avg_log2FC, pct.1, pct.2)
  top_markers_tbl

  marker_plot_df <- markers %>%
    filter(!is.na(p_val_adj), !is.na(avg_log2FC)) %>%
    mutate(
      cluster = factor(cluster),
      neg_log10_padj = -log10(pmax(p_val_adj, 1e-300))
    )

  label_df <- marker_plot_df %>%
    group_by(cluster) %>%
    slice_max(order_by = avg_log2FC, n = 2, with_ties = FALSE) %>%
    ungroup()

  ggplot(marker_plot_df, aes(x = avg_log2FC, y = neg_log10_padj)) +
    geom_point(alpha = 0.35, size = 0.7, color = "#238b45") +
    facet_wrap(~ cluster, scales = "free_y") +
    geom_text(
      data = label_df,
      aes(label = gene),
      size = 2.4,
      vjust = -0.5,
      check_overlap = TRUE
    ) +
    labs(
      title = "DESeq2 marker significance by cluster",
      x = "avg_log2FC",
      y = "-log10(adjusted p-value)"
    ) +
    theme_bw()
} else {
  message("Marker discovery skipped. Set RUN_MARKERS <- TRUE to run it and generate DESeq2 marker plots.")
}
```

# Gene boxplots

**Why these genes:**

- `EPCAM` and `KRT19` flag epithelial programs.
- `VIM` points to mesenchymal-like states.
- `MKI67` and `TOP2A` highlight cycling / proliferative clusters.
- `ERBB2` and `ESR1` map onto HER2/ER signaling that’s central in breast cancer.

```{r gene-boxplots, fig.width=11, fig.height=18}
plot_genes_sym <- symbol_sets$PlotGenes
plot_gene_labels <- setNames(plot_genes_sym, symbol_to_ens[plot_genes_sym])
plot_gene_labels <- plot_gene_labels[!is.na(names(plot_gene_labels))]
plot_genes <- names(plot_gene_labels)

expr <- FetchData(bc, vars = c("seurat_clusters", plot_genes))
expr_long <- expr %>%
  pivot_longer(cols = all_of(plot_genes), names_to = "gene", values_to = "expr")

expr_long$seurat_clusters <- factor(expr_long$seurat_clusters)
expr_long$gene <- plot_gene_labels[expr_long$gene]

plot_list <- lapply(unique(expr_long$gene), function(g) {
  ggplot(filter(expr_long, gene == g), aes(x = seurat_clusters, y = expr, fill = seurat_clusters)) +
    geom_boxplot(outlier.shape = NA, alpha = 0.85) +
    geom_jitter(size = 0.2, width = 0.2, alpha = 0.25) +
    scale_fill_viridis_d() +
    labs(title = g, x = "Cluster", y = "Expression") +
    theme_minimal(base_size = 11)
})

wrap_plots(plot_list, ncol = 1)
```

# Quick validation checks

These are lightweight checks to make sure the structure isn’t driven by QC artifacts.

```{r validation-qc, fig.width=11, fig.height=4}
FeatureScatter(bc, feature1 = "nCount_RNA", feature2 = "percent.mt") +
  FeatureScatter(bc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
```

```{r validation-umap, fig.width=9, fig.height=6}
FeaturePlot(bc, features = c("nCount_RNA", "percent.mt"), ncol = 2)
```

```{r validation-cluster-qc, fig.width=10, fig.height=5}
VlnPlot(bc, features = c("nCount_RNA", "percent.mt"), group.by = "seurat_clusters", ncol = 2)
```

# What else you can do next

- Integrate with bulk RNA-seq or drug response data to connect expression states to sensitivity.
- Run pathway enrichment on each cluster for cleaner biological narratives.
- Compare these cell-line states to patient tumor single-cell datasets to see what carries over.

# Summary

This analysis maps single-cell structure across 32 breast cancer cell lines, highlights major expression states (epithelial, mesenchymal, proliferative, stress), and backs the cluster interpretation with markers and targeted gene plots. It’s a compact example of how to go from raw counts to a story you can actually explain.
